---
title: "Problem Set 4, Econ 211C"
author: "Swati Sharma"
output:
  pdf_document: default
  html_document: default
  html_notebook: default
---
```{r warning=F, message =F, echo = F}
library(Quandl)
library(DataCombine)

```

# Question 1 (40 points)

Consider a special case of the long-run risks model:
\begin{gather}
x_{t+1} = \rho x_{t} + \varphi_{e}\sigma e_{t+1} \\
g_{t+1} = \log\left(C_{t+1}/C_{t}\right) = \mu + x_{t} + \sigma
\eta_{t+1} \\
e_{t+1},\eta_{t+1} \overset{i.i.d.}{\sim} N(0,1) \\
\end{gather}
$\mu = 0.0015$, $\sigma = 0.0078$, $\rho = 0.979$, $\varphi_e = 0.044$, and $\varphi_d = 4.5$. Download real personal consumption expenditures data from the BEA for the period Jan 1999 -- Mar 2018 and use the sum of non-durable and services as a monthly series for consumption (note: the Quandl code for the data is `BEA/T20806_M`). Estimate values for the latent state $\hat{\xi}_{t+1|t} = \hat{x}_{t+1|t}$ for each month using the Kalman filter. Write the recursions yourself (without using any `R` packages) and plot the forecasts of the latent states.

#### Solution:

```{r}
consumption = Quandl("BEA/T20806_M",start_date="1999-01-01",end_date="2018-03-01",type="xts")
temp = as.data.frame(consumption)
```

```{r}
temp$consumption = temp$`:Nondurable goods` + temp$Services

temp <- slide(temp, Var = "consumption", slideBy = -1)

temp$g_t = log(temp$consumption/temp$`consumption-1`)

temp <- temp[,-c(1:10)]

g_t = temp$g_t

```

```{r}
xhat = rep(0, nrow(temp))
ghat = rep(0, nrow(temp))
P = rep(0, nrow(temp))

epsilon <- rnorm(nrow(temp),0,1)
eta <- rnorm(nrow(temp),0,1)

mu = 0.0015
sigma = 0.0078
rho = 0.979
phi_e = 0.044
phi_d = 4.5

# initializing guess 
xhat_init <- 0
P_init <- 1
H <- 1
A <- mu
R <- sigma^2
x_t <- 1
F <- rho
Q <- (phi_e^2)*(sigma^2)
r = sigma*eta*phi_e
u = sigma*eta


S = rep(0, nrow(temp))
S[1] = H*P_init*H+R

xhat[1] = xhat_init + r[2]+ P_init*H*(S[1])*(g_t[2]-A*x_t+H*xhat_init-u[2])
P[1]= P_init - P_init*H*(S[1])*H*P_init

for(i in 1: nrow(temp)) { 
xhat[i+1] = F*xhat_init+ r[i+1]+ F*P_init*H*(S[i])*(g_t[i+1]-A*x_t+H*xhat[i]-u[2])
P[i+1] = F*(P_init-P_init*H*(S[i])*H*P_init)*F+Q

ghat[i+1] = A*x_t+H*xhat[i+1]+r[i+2]
S[i+1] = H*P[i+1]*H+R 
}

plot(ghat, main = "Plot of Forecast of Latent States", type = "l", xlab = "Time")
```

# Question 2 (30 points)

Suppose that $r_1$, $r_2$, \ldots, $r_n$ are observations of a return series that follows the $AR(1)-GARCH(1,1)$ model:
\begin{align}
  r_t & = \mu + \phi r_{t-1} + \sigma_t \varepsilon_t \\
  \sigma_t^2 & = \alpha_0 + \alpha_1 \sigma_{t-1}^2 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2,
\end{align}
where $\varepsilon_t \stackrel{i.i.d.}{\sim} N(0,1)$. Derive the conditional log likelihood of the data.

#### Solution:
Given the above equations we can determine that 
\[E(r_t)=\mu+\phi_{t-1}\]
\[ML(\theta|y_1)=\prod f_{Y_t|Y_{t-1}}\]
where
\[f_{Y_t|Y_{t-1}}=N(\mu+\phi_{t-1},\alpha_0 + \alpha_1 \sigma_{t-1}^2 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2)\]
\[ML(\theta|y_1)=\prod \frac{1}{\sqrt{2\pi*(\alpha_0 + \alpha_1 \sigma_{t-1}^2 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2)}}*exp(\frac{-(r_t-\mu-\phi r_{t-1})^2}{2*(\alpha_0 + \alpha_1 \sigma_{t-1}^2 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2))})\]
\[ML(\theta|y_t)=\sum_{t=2}^T log(f_{Y_t|Y_{t-1}})\]
\[ML(\theta|y_t)=\sum_{t=2}^T=-\frac{T-1}{2}log(2\pi)-\frac{1}{2}\sum_{t=2}^T log(\alpha_0 + \alpha_1 \sigma_{t-1}^2 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2)-\sum_{t=2}^T\frac{(r_t-\mu-\phi r_{t-1})^2}{2*(\alpha_0 + \alpha_1 \sigma_{t-1}^2 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2))}\]


# Question 3 (30 points)

Consider the monthly returns of Intel stock from January 1973 to May 2018. Transform the returns into log returns. Build a GARCH model for the transformed series and compute 1-step to 5-step ahead volatility forecasts at the forecast origin May 2018. Plot your volatility estimates for the entire period, along with your forecasts.

#### Solution:
```{r}
library(quantmod)
library(dplyr)
getSymbols("INTC", from = "1973-01-01", to = "2018-05-31")
monthRets <- as.data.frame(monthlyReturn(log(INTC$INTC.Adjusted)))
monthRets <- tibble::rownames_to_column(monthRets)
```

```{r}
library(fGarch)
library(ggplot2)
library(lubridate)
library(ggthemes)
## Finding step 1:5 Volatility forecasts
garch.1 <- garchFit(formula = ~garch(1,1), data = monthRets$monthly.returns, trace = FALSE)
volGARCH <- as.data.frame(predict(garch.1, 5)$standardDeviation)
volGARCH$dates <- c("2018-06-30", "2018-07-31", "2018-08-31", "2018-09-29", "2018-10-31")

## Pulling estimated historic volatility
volMonthlyRets <- as.data.frame(volatility(garch.1,type = "sigma"))
volMonthlyRets$dates <- monthRets$rowname
```

```{r}
ggplot()+
  geom_line(data=volMonthlyRets, aes(y=`volatility(garch.1, type = "sigma")`, x=as.Date(dates), linetype="Historic"))+
  geom_line(data=volGARCH, aes(y=`predict(garch.1, 5)$standardDeviation`, x=as.Date(dates), linetype="Forecasted"), colour="blue")+
  scale_linetype_manual(name="", values = c("Historic"=3, "Forecasted"=1))+
  ylab("Volatility")+
  xlab("Dates")+
  ggtitle("Historic and Forecasted Volatility of Monthly Log Returns of Intel")+
  theme_tufte()


```


